\section{Numerical Stability and Precision}
\label{sec:numerical-stability}

\subsection{Classical diagonalization}

All classical eigenvalue computations were performed using \texttt{numpy.linalg.eigh}, which calls the LAPACK \texttt{dsyevd} routine operating in IEEE 754 double-precision floating point (64-bit). For the two-level Hamiltonian the maximum discrepancy between the analytical eigenvalues and the numerical ones is $\approx10^{-15}$, which is consistent with the machine epsilon $\epsilon_\text{mach}\approx2.2\times10^{-16}$ and reflects a single rounding-error accumulation from the few arithmetic operations involved. No loss of precision was observed across the full $\lambda\in[0,1]$ sweep, including near the avoided crossing at $\lambda=\frac{2}{3}$, where the gap is $\Delta E\approx0.27$ — well above the numerical noise floor.

For the Lipkin matrices ($3\times3$ and $5\times5$), diagonalization remains numerically trivial. The condition numbers of these small Hermitian matrices stay moderate even at the largest interaction strengths studied, and no ill-conditioning effects were detected.

\subsection{VQE precision and optimizer behavior}

Since all VQE experiments use exact statevector simulation rather than finite-shot sampling, the dominant sources of numerical error are the optimizer convergence tolerance and the expressibility of the chosen ansatz relative to the true ground state. \Cref{tab:vqe-precision-summary} summarizes the achieved precision across all four VQE experiments.

\begin{table*}[t]
\centering
\begin{tabular}{lcccc}
\hline
System & Ansatz & Optimizer & Max error & Mean error \\
\hline
$2\times2$ model (Part c)      & $R_y$                        & BFGS     & $5.4\times10^{-11}$ & $\sim10^{-12}$ \\
Two-qubit model (Part e)       & $R_y$-CNOT-$R_y$             & BFGS     & $\sim10^{-10}$      & $\sim10^{-12}$ \\
Lipkin $J=1$ (Part g)          & $R_y$-CZ-$R_y$, depth 2      & L-BFGS-B & $\sim10^{-11}$      & $\sim10^{-12}$ \\
Lipkin $J=2$ (Part g)          & $R_y$-CZ-$R_y$, depth 4      & L-BFGS-B & $\sim10^{-2}$       & $\sim10^{-3}$  \\
\hline
\end{tabular}
\caption{Summary of VQE precision across all systems. Errors are absolute differences between the VQE ground state energy and the exact classical result, evaluated over the full parameter sweep.}
\label{tab:vqe-precision-summary}
\end{table*}

For the single-qubit $2\times2$ Hamiltonian, the $R_y$ ansatz spans the full real single-qubit Hilbert space, so the ansatz is exact by construction. The residual error of $\approx5.4\times10^{-11}$ reflects the BFGS convergence tolerance rather than any approximation in the trial state. The one-parameter energy landscape $E(\theta)$ is smooth and unimodal for all $\lambda$, so no local minima exist and a single optimization from $\theta=0$ reliably reaches the global minimum.

For the two-qubit model, the $R_y$-CNOT-$R_y$ circuit with four parameters spans the full real two-qubit Hilbert space. The optimization landscape is again free of local minima for this system, and BFGS converges from the zero initialization $\boldsymbol{\theta}=(0,0,0,0)$ without requiring random restarts. The achieved precision of $\approx10^{-10}$ is again optimizer-limited.

For the Lipkin $J=1$ system, the two-qubit hardware-efficient ansatz achieves similar machine-level precision. The block structure of the Lipkin Hamiltonian and the low dimensionality of the $J=1$ sector produce a landscape that L-BFGS-B navigates reliably with three random restarts.

The Lipkin $J=2$ system is the most demanding. The $16\times16$ qubit Hilbert space contains $J=0,1,2$ sectors; the ansatz must implicitly project onto the five-dimensional $J=2$ subspace. With depth 4 and 16 parameters, the circuit has sufficient expressibility in principle, but the optimization landscape is more complex. A residual error of $\approx10^{-2}$ was observed, indicating that the optimizer converged to a near-minimum that is not the global one. Four random restarts partially mitigated this, suggesting a rugged landscape with multiple local minima in the $J=2$ case.

\subsection{Shot noise}

Although the main experiments use exact statevector simulation, the Bell state measurement experiment (Part a) was performed with finite shots ($1000$ shots, $50$ repetitions). The expected shot noise for a balanced outcome ($p=0.5$) is
\begin{equation}
    \sigma_\text{shot} = \sqrt{\frac{p(1-p)}{N_\text{shots}}} = \sqrt{\frac{0.5\times0.5}{1000}} \approx 0.0158,
\end{equation}
in agreement with the observed standard deviation of $\sigma\approx0.018$. This confirms that the measurement noise is purely statistical and consistent with binomial sampling.

\subsection{Reproducibility}

All random restarts in the VQE experiments use a fixed random seed (\texttt{42}) for reproducibility, with random initial parameters drawn uniformly from $[-\pi,\pi]$. Default \texttt{scipy.optimize} tolerances were used throughout; no explicit convergence thresholds were set beyond the optimizer defaults. No catastrophic cancellation or overflow was encountered in any of the experiments, and all results were verified to be stable under repeated runs.
